Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-09 method.tex
Converted to AASTeX
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@997 ccdd05d5-56c1-461a-bf98-85f6876523b4

\section{Early warning searches for compact binary coalescence}
\label{SECII}\label{sec:method}

In this section we describe a decomposition of the \CBC{} waveform space that
reduces time-domain (\TD) filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multi-band decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR3 joint science
runs~\cite{HugheyGWPAW2011}.  We combine this with the \SVD\ rank-reduction
method of~\cite{Cannon:2010p10398} that exploits the redundancy of
the template banks.

@@

Inspiral signals are continuously parameterized by a set of intrinsic source
parameters $\vec\theta$ that determine the amplitude and phase evolution of the
\GW{} strain. For systems where the effects of spin can be
ignored, the intrinsic source parameters are the component masses of the
binary, $\vec\theta = (m_1, m_2)$. Searches for inspiral signals typically
employ matched filter banks that discretely sample the possible intrinsic
parameters~\cite{findchirppaper}.  For a given source, the strain observed by
the detector is a linear combination of two waveforms corresponding to the
`$+$' and `$\times$' \GW{} polarizations.  Thus, for any value of
$\vec\theta$ we must implement 2 filters.  The coefficients for the $\numtmps$ filters
are known as templates, and are formed by discretizing and time reversing the
waveforms and weighting them by the inverse amplitude spectral density of the
detector's noise. To construct a template bank, templates are chosen with the
$\numtmps/2$ discrete signal parameters $\vec\theta_0,\, \vec\theta_1,\, \dots,\,
\vec\theta_{\numtmps/2-1}$ to assure a bounded loss of \SNR~\cite{Owen:1998dk}. That
is, any possible signal within a given range of intrinsic
source parameters will have an inner product that is $\geqslant 0.97$ with at
least one template. Such a template bank is said to have a {\em minimum match}
of 0.97. The data from the detector are whitened and convolved with each
template to produce $\numtmps$ \SNR\ time series. Local peak-finding across time and
templates determines detection candidates.

@@

The first transformation is to chop the \TD{} templates up into \emph{time
slices}.  Since each template slice is disjoint in time, the resulting set for
a single template is orthogonal.  Given the chirp-like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely down-sampled.  The down-sampling reduces the number of data samples by a
factor of $\sim$100 from the early part of the waveform and allows the filters
to be evaluated at about $\sim$$1/100^\mathrm{th}$ the sample rate. This
amounts to more than a factor of 10000
reduction in the floating-point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the \SVD{} to produce an orthogonal filter set
from the time-sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of
$\sim$100.  The combined methods reduce the number of floating point operations
to the level where they are competitive with the conventional high-latency \FD\
matched filter approach.  In the remainder of this section we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

The first step of our proposed method is to divide the templates into time
slices in a \TD{} analogue to the \FD{} decomposition described
in~\cite{Marion2004, Buskulic2010}.  To wit,
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\cases{%
		h_i^s[k] & for $t^s \leqslant k / f^0 < t^{s+1}$ \\
		0 & otherwise
	}
\end{equation}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0
t^S = N$.  A matched filter is constructed for each time slice.  The outputs
form an ensemble of partial \SNR{} streams.  By linearity, these partial \SNR\
streams can be suitably time-delayed and summed to reproduce the \SNR\ of the
full template.  We will show in the next section that this, combined with the
\SVD{}, is sufficient to enable a computationally efficient \TD{} search and
furthermore is an essential part of an \earlywarning\ detection scheme.

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrapole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \umu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t^0$, or equivalently frequency
$f_\mathrm{hi}$.  This is often chosen to correspond to the \ISCO. An inspiral
signal will enter the detection band at a low frequency, $f = f_\mathrm{low}$,
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t^0, t_\mathrm{low})$ and is said to have  a duration of
$t_\mathrm{low} - t^0$. It is critically sampled at a rate of $2
f_\mathrm{hi}$.

The monotonic time-frequency relationship of equation~\eqref{eq:fgw} allows us
to choose time-slice boundaries that require substantially less bandwidth at
early times in the inspiral.  Our goal is to reduce the filtering cost of a
large fraction of the waveform by computing part of the convolution at a lower
sample rate.  Specifically we consider here time slice boundaries with the
smallest power-of-two sample rate that sub-critically samples the time-sliced
template.  The time slices consist of the $S$ intervals
$\left[t^0, t^1\right),\, \left[t^1, t^2\right),\, \dots,\, \left[t^{S-1}, t^S\right)$,
sampled at frequencies $f^0,\, f^1,\, \dots,\, f^\mathrm{S-1}$ where $f^0
\geqslant 2 f_\mathrm{hi}$ and $f^{S-1} \geqslant 2 f_\mathrm{low}$.  The time
sliced templates may be down-sampled without aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\cases{
		h_{i}\!\left[k\frac{f}{f^s}\right] & if $t^s \leqslant k/f^s < t^{s+1}$ \\
		0 & otherwise.
	}
\end{equation}
%
\begin{comment}
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \,
M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{hi} = f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time-frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceding coalescence and \slicessamps\ is the number of samples in the
$s^{\mathrm{th}}$ filter.}
\end{table}
\end{comment}

Since waveforms with neighboring intrinsic source parameters $\vec\theta$ have
similar time-frequency evolution, it is possible to design computationally
efficient time slices for an extended region of parameter space rather than to
design different time slices for each template.

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of \SVD\ {\em
basis templates}, previously demonstrated
in~\cite{Cannon:2010p10398}.  Similarly, the time-sliced templates described
above can be approximated to arbitrary accuracy by expansion into a set of
\SVD\ \emph{basis templates}, $u_l^s[k]$, related to the original time
sliced templates through the \emph{reconstruction matrix},
$v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
The parameter $\numsvdtmps$ is the number of basis templates that are kept in
the approximation.  This determines the \SVD\ tolerance, which affects the
\SNR\ loss due to the approximation.  The authors of \cite{Cannon:2010p10398}
showed that high accuracy could be achieved with far fewer basis templates than
templates in the original template bank.  We find that when combined with the
time-slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps\ and improves on
\cite{Cannon:2010p10398} by nearly an order of magnitude.  In the next section
we describe how we form our early-warning detection statistic using the time-slice
decomposition and the \SVD.

@@

In the previous two sections we have described two transformations that greatly
reduce the burden of filtering waveforms from the the compact binary parameter space and make \TD\
filtering of the data possible.  We are now prepared to define the early
warning \SNR\ and to comment on the computational cost of evaluating it.  But
first, we will introduce some notation referring to the decimation of data,
which means reducing the sample rate after low-pass filtering,
\begin{equation}
\label{eq:decomp}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k] \,,
\end{equation}
and notation referring to the interpolation of data to increase the sample rate,
\begin{equation}
	x^{s}[k] = \left( H^\shortuparrow x^{s+1}\right)[k] \,.
\end{equation}

From the combination of transformations to the compact binary templates defined
in equation \eqref{eq:time-sliced-templates} and \eqref{eq:svddecomp} we define
the early-warning filter output accumulated up to time slice $s$ as,
%
% orthogonal decomposition filtering
%
\begin{equation}
	\rho_i^s [k] =%
		% Reconstruction
		\underbrace{
			\sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s
		}_\textrm{\clap{reconstruction}}
		% Orthogonal FIR filter
		\overbrace{
			\sum_{\mathclap{n=0}}^{\mathclap{N^s-1}} u_l^s[n] x^s[k-n]
		}^\textrm{\clap{orthogonal {\sc fir} filters}}
		% Plus ...
		+
		% Interpolation SNR
		\underbrace{
			\left(H^\uparrow \rho_i^{s+1}\right)[k]
		}_\textrm{\clap{{\sc snr} from previous time slices}}
\end{equation}
%
%
This quantity is related to the early-warning \SNR{} by the normalization
of the partial filter defined by the present time slice, which may be computed
in advance. \editorial{Chad: are we happy with that statement?}
%
%
\begin{figure}[htbp]
	\includegraphics{figures/lloid-diagram.pdf}
	\caption{\label{fig:pipeline} Schematic of \lloid{} pipeline illustrating
signal flow.  Circles with arrows represent interpolation
\protect\includegraphics{figures/upsample-symbol.pdf} or decimation
\protect\includegraphics{figures/downsample-symbol.pdf}.  Circles with plus
signs represent summing junctions
\protect\includegraphics{figures/adder-symbol.pdf}.  Squares
\protect\includegraphics{figures/fir-symbol.pdf} stand for \fir{} filters.  Sample
rate decreases from the top of the diagram to the bottom.  In this diagram each
time slice contains three \fir\ filters that are linearly combined to produce
four output channels.  In a typical pipeline the number of \fir\ filters is
much less than the number of output channels.}
\end{figure}
%
%
The signal flow diagram in figure~\ref{fig:pipeline} illustrates how this
recursion relation can be realized as a multi-rate filter network with outputs for
each of the early-warning \SNR{}s.

@@

We now examine the computational cost scaling of the conventional \TD{} or
\FD{} matched filter procedure as compared with \lloid{}.  For convenience,
table~\ref{tab:recap} provides a review of the notation that we will need in
this section.
%
% symbols used in FLOPs calculations table
%
\Table{\label{tab:recap} Notation used to describe filters.}
\br
	& Definition \\
\mr
\numtmps		& number of templates \\
\tmpsamps	& number of samples per template \\
\numslices	& number of time slices \\
\numsvdtmps	& number of orthogonal templates in time slice $s$ \\
\slicessamps	& number of samples in decimated time slice $s$\\
$f^s$		& sample rate in time slice $s$ \\
$N^\shortdownarrow$ & number of coefficients in decimation filter \\
$N^\shortuparrow$ & number of coefficients in interpolation filter \\
\br
\end{tabular}
\end{indented}
\end{table}

\editorial{This section's two tables are small.  Any way to put them side by side or next to another figure?}

@@

The most common \FD{} method is known as the \emph{overlap-save} algorithm, described in \cite{numerical-recipes-chapter-13}.  It entails splitting the input into blocks of $D$ samples, $D > \tmpsamps$, each block overlapping the previous one by $D - \tmpsamps$ samples.  For each block, the algorithm computes the forward \fft\ of the data and the templates, multiplies them, and then computes the reverse \fft.

Modern implementations of the \fft, such as the ubiquitous \texttt{fftw}, require about $2 \fftblock \lg \fftblock$ operations to evaluate a real transform of size $\fftblock$~\cite{Johnson:2007p9654}.  Including the forward transform of the data and $M$ reverse transforms for all of the templates, the \fft\ costs $2 (\numtmps + 1) \fftblock \lg \fftblock$ operations per block.  The multiplication of the transforms adds a further $2 \numtmps \fftblock$ operations per block.  Since each block produces $\fftblock - \tmpsamps$ usable samples of output, the overlap-save method requires
$$
f^0 \cdot \frac{2 (\numtmps + 1) \lg \fftblock + 2 \numtmps}{1 - \tmpsamps/\fftblock} \; \mathrm{\flops} \,.
$$

>>>>

\section{Early warning searches for compact binary coalescence}
\label{sec:method}

In this section we describe a decomposition of the \CBC{} waveform space that
reduces time-domain (\TD) filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\citep{Marion2004, Buskulic2010} that describe a
multi-band decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR3 joint science
runs~\citep{HugheyGWPAW2011}.  We combine this with the \SVD\ rank-reduction
method of~\citep{Cannon:2010p10398} that exploits the redundancy of
the template banks.

@@

Inspiral signals are continuously parameterized by a set of intrinsic source
parameters $\mathbf\Theta$ that determine the amplitude and phase evolution of the
\GW{} strain. For systems where the effects of spin can be
ignored, the intrinsic source parameters are the component masses of the
binary, $\mathbf\Theta = (m_1, m_2)$. Searches for inspiral signals typically
employ matched filter banks that discretely sample the possible intrinsic
parameters~\citep{findchirppaper}.  For a given source, the strain observed by
the detector is a linear combination of two waveforms corresponding to the
`$+$' and `$\times$' \GW{} polarizations.  Thus, for any value of
$\mathbf\Theta$ we must implement 2 filters.  The coefficients for the $\numtmps$ filters
are known as templates, and are formed by discretizing and time reversing the
waveforms and weighting them by the inverse amplitude spectral density of the
detector's noise. To construct a template bank, templates are chosen with the
$\numtmps/2$ discrete signal parameters $\mathbf\Theta_0,\, \mathbf\Theta_1,\, \dots,\,
\mathbf\Theta_{\numtmps/2-1}$ to assure a bounded loss of \SNR~\citep{Owen:1998dk}. That
is, any possible signal within a given range of intrinsic
source parameters will have an inner product that is $\geqslant 0.97$ with at
least one template. Such a template bank is said to have a {\em minimum match}
of 0.97. The data from the detector are whitened and convolved with each
template to produce $\numtmps$ \SNR\ time series. Local peak-finding across time and
templates determines detection candidates.

@@

The first transformation is to chop the \TD{} templates up into \emph{time
slices}.  Since each template slice is disjoint in time, the resulting set for
a single template is orthogonal.  Given the chirp-like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely down-sampled.  The down-sampling reduces the number of data samples by a
factor of $\sim$100 from the early part of the waveform and allows the filters
to be evaluated at about $\sim$$1/100^\mathrm{th}$ the sample rate. This
amounts to more than a factor of 10000
reduction in the floating-point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the \SVD{} to produce an orthogonal filter set
from the time-sliced templates~\citep{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of
$\sim$100.  The combined methods reduce the number of floating point operations
to the level where they are competitive with the conventional high-latency \FD\
matched filter approach.  In the remainder of this section we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

The first step of our proposed method is to divide the templates into time
slices in a \TD{} analogue to the \FD{} decomposition described
in~\citep{Marion2004, Buskulic2010}.  To wit,
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\begin{cases}
		h_i^s[k] & \textrm{if } t^s \leqslant k / f^0 < t^{s+1} \\
		0 & \textrm{otherwise}
	\end{cases}
\end{equation}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0
t^S = N$.  A matched filter is constructed for each time slice.  The outputs
form an ensemble of partial \SNR{} streams.  By linearity, these partial \SNR\
streams can be suitably time-delayed and summed to reproduce the \SNR\ of the
full template.  We will show in the next section that this, combined with the
\SVD{}, is sufficient to enable a computationally efficient \TD{} search and
furthermore is an essential part of an \earlywarning\ detection scheme.

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrapole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \upmu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\citep{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t^0$, or equivalently frequency
$f_\mathrm{hi}$.  This is often chosen to correspond to the \ISCO. An inspiral
signal will enter the detection band at a low frequency, $f = f_\mathrm{low}$,
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t^0, t_\mathrm{low})$ and is said to have  a duration of
$t_\mathrm{low} - t^0$. It is critically sampled at a rate of $2
f_\mathrm{hi}$.

The monotonic time-frequency relationship of equation~\eqref{eq:fgw} allows us
to choose time-slice boundaries that require substantially less bandwidth at
early times in the inspiral.  Our goal is to reduce the filtering cost of a
large fraction of the waveform by computing part of the convolution at a lower
sample rate.  Specifically we consider here time slice boundaries with the
smallest power-of-two sample rate that sub-critically samples the time-sliced
template.  The time slices consist of the $S$ intervals
$\left[t^0, t^1\right),\, \left[t^1, t^2\right),\, \dots,\, \left[t^{S-1}, t^S\right)$,
sampled at frequencies $f^0,\, f^1,\, \dots,\, f^\mathrm{S-1}$ where $f^0
\geqslant 2 f_\mathrm{hi}$ and $f^{S-1} \geqslant 2 f_\mathrm{low}$.  The time
sliced templates may be down-sampled without aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\begin{cases}
		h_{i}\!\left[k\frac{f}{f^s}\right] & \textrm{if } t^s \leqslant k/f^s < t^{s+1} \\
		0 & \textrm{otherwise.}
	\end{cases}
\end{equation}
%
\begin{comment}
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \,
M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{hi} = f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time-frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceding coalescence and \slicessamps\ is the number of samples in the
$s^{\mathrm{th}}$ filter.}
\end{table}
\end{comment}

Since waveforms with neighboring intrinsic source parameters $\mathbf\Theta$ have
similar time-frequency evolution, it is possible to design computationally
efficient time slices for an extended region of parameter space rather than to
design different time slices for each template.

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of \SVD\ {\em
basis templates}, previously demonstrated
in~\citep{Cannon:2010p10398}.  Similarly, the time-sliced templates described
above can be approximated to arbitrary accuracy by expansion into a set of
\SVD\ \emph{basis templates}, $u_l^s[k]$, related to the original time
sliced templates through the \emph{reconstruction matrix},
$v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
The parameter $\numsvdtmps$ is the number of basis templates that are kept in
the approximation.  This determines the \SVD\ tolerance, which affects the
\SNR\ loss due to the approximation.  The authors of \citep{Cannon:2010p10398}
showed that high accuracy could be achieved with far fewer basis templates than
templates in the original template bank.  We find that when combined with the
time-slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps\ and improves on
\citep{Cannon:2010p10398} by nearly an order of magnitude.  In the next section
we describe how we form our early-warning detection statistic using the time-slice
decomposition and the \SVD.

@@

In the previous two sections we have described two transformations that greatly
reduce the burden of filtering waveforms from the the compact binary parameter space and make \TD\
filtering of the data possible.  We are now prepared to define the early
warning \SNR\ and to comment on the computational cost of evaluating it.  But
first, we will introduce some notation referring to the decimation of data,
which means reducing the sample rate after low-pass filtering,
\begin{equation*}
\label{eq:decomp}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k] \,,
\end{equation*}
and notation referring to the interpolation of data to increase the sample rate,
\begin{equation*}
	x^{s}[k] = \left( H^\shortuparrow x^{s+1}\right)[k] \,.
\end{equation*}

From the combination of transformations to the compact binary templates defined
in equation \eqref{eq:time-sliced-templates} and \eqref{eq:svddecomp} we define
the early-warning filter output accumulated up to time slice $s$ as,
%
% orthogonal decomposition filtering
%
\begin{equation}
	\rho_i^s [k] =%
		% Reconstruction
		\underbrace{
			\sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s
		}_\textrm{\clap{reconstruction}}
		% Orthogonal FIR filter
		\overbrace{
			\sum_{\mathclap{n=0}}^{\mathclap{N^s-1}} u_l^s[n] x^s[k-n]
		}^\textrm{\clap{orthogonal {\sc fir} filters}}
		% Plus ...
		+
		% Interpolation SNR
		\underbrace{
			\left(H^\uparrow \rho_i^{s+1}\right)[k]
		}_\textrm{\clap{{\sc snr} from previous time slices}}
\end{equation}
%
%
This quantity is related to the early-warning \SNR{} by the normalization
of the partial filter defined by the present time slice, which may be computed
in advance. \editorial{Chad: are we happy with that statement?}
%
%
\begin{figure}[h]
	\plotone{figures/lloid-diagram.pdf}
	\caption{\label{fig:pipeline} Schematic of \lloid{} pipeline illustrating
signal flow.  Circles with arrows represent interpolation
\protect\includegraphics{figures/upsample-symbol.pdf} or decimation
\protect\includegraphics{figures/downsample-symbol.pdf}.  Circles with plus
signs represent summing junctions
\protect\includegraphics{figures/adder-symbol.pdf}.  Squares
\protect\includegraphics{figures/fir-symbol.pdf} stand for \fir{} filters.  Sample
rate decreases from the top of the diagram to the bottom.  In this diagram each
time slice contains three \fir\ filters that are linearly combined to produce
four output channels.  In a typical pipeline the number of \fir\ filters is
much less than the number of output channels.}
\end{figure}
%
%
The signal flow diagram in figure~\ref{fig:pipeline} illustrates how this
recursion relation can be realized as a multi-rate filter network with outputs for
each of the early-warning \SNR{}s.

@@

We now examine the computational cost scaling of the conventional \TD{} or
\FD{} matched filter procedure as compared with \lloid{}.  For convenience,
table~\ref{tab:recap} provides a review of the notation that we will need in
this section.
%
% symbols used in FLOPs calculations table
%
\begin{table}
\caption{\label{tab:recap}Notation used to describe filters.}
\begin{tabular}{ll}
\tableline\tableline
& Definition \\
\tableline
\numtmps		& number of templates \\
\tmpsamps	& number of samples per template \\
\numslices	& number of time slices \\
\numsvdtmps	& number of orthogonal templates in time slice $s$ \\
\slicessamps	& number of samples in decimated time slice $s$\\
$f^s$		& sample rate in time slice $s$ \\
$N^\shortdownarrow$ & number of coefficients in decimation filter \\
$N^\shortuparrow$ & number of coefficients in interpolation filter \\
\tableline
\end{tabular}
\end{table}

@@

The most common \FD{} method is known as the \emph{overlap-save} algorithm, described in \citep{numerical-recipes-chapter-13}.  It entails splitting the input into blocks of $D$ samples, $D > \tmpsamps$, each block overlapping the previous one by $D - \tmpsamps$ samples.  For each block, the algorithm computes the forward \fft\ of the data and the templates, multiplies them, and then computes the reverse \fft.

Modern implementations of the \fft, such as the ubiquitous \texttt{fftw}, require about $2 \fftblock \lg \fftblock$ operations to evaluate a real transform of size $\fftblock$~\citep{Johnson:2007p9654}.  Including the forward transform of the data and $M$ reverse transforms for all of the templates, the \fft\ costs $2 (\numtmps + 1) \fftblock \lg \fftblock$ operations per block.  The multiplication of the transforms adds a further $2 \numtmps \fftblock$ operations per block.  Since each block produces $\fftblock - \tmpsamps$ usable samples of output, the overlap-save method requires
$$
f^0 \cdot \frac{2 (\numtmps + 1) \lg \fftblock + 2 \numtmps}{1 - \tmpsamps/\fftblock} \; \mathrm{\flops} \,.
$$
